\addchap{Acknowledgments}

First and foremost, I want to thank Ulrike Sayatz, without whom the work presented here would not exist.
Ulrike convinced me that graphemics is a highly relevant discipline and an integral and indispensable part of linguistics and grammar.
Together, we developed the ideas of \textit{usage-based graphemics}, which underlie some of the work presented here.
Ulrike also helped me to understand German morphosyntax much better through many discussions.
Most importantly, she has been a productive and inspiring co-author for many years now.

I also thank Elizabeth Pankratz for being an inspiring collaborator, co-author and moral supporter.
Among other things, Elizabeth changed the way I look at compounds, leading to significant improvements of the DECOW corpus and joint research on compounds.

Furthermore, I am grateful to Felix Bildhauer for ongoing discussions about corpus construction, corpus analysis, and statistics (worth a substantial amount of Złoty).
Without his collaboration, the COW corpora would not exist, and I could not have conducted the research presented in this book.

For their support as teachers, colleagues, employers, enablers, thesis advisors, and moral supporters (or any combination thereof), I am indebted to (in alphabetical order) Dirk Buschbom, Susanne Flach, Thomas M.\ Groß, Iris Hasselberg, Götz Keydana, Michael Job, Stefan Müller, Bjarne Ørsnes, Erich Poppe, Manfred Sailer, and Gert Webelhuth.

Student assistants who contributed significantly to the success of my research by doing a lot of painstaking work on corpus data and supervising experiments are (in alphabetical order) Sarah Dietzfelbinger, Lea Helmers, Kim Maser, and Luise Rißmann.
I am very grateful to them.
Also, I want to thank a few students who attended my classes at Freie Universität Berlin between 2008 and 2017, and who managed to inspire my work by asking intriguing questions and initiating stimulating discussions: Johanna Rehak, Samuel Reichert, and Nicolai Sinn (in alphabetical order).

Finally, I want to thank everyone who supported me on a personal level over the years, including but surely not restricted to (in alphabetical order) Matthias B.\ Döring, Michael Karg, Tanja Hagedorn, Julia Schmidt, and my parents.

%My work on this project was funded in part by the \textit{Deutsche Forschungsgemeinschaft} (DFG, personal grant SCHA1916/1-1) through the project \textit{Linguistic Web Characterisation}.
